Visualizing Contingency Tables

### 2006-04-25 trunk/vcd/paper/hsc.Rnw
revised section 4 in HSC chapter
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@524 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

In the following, we will first treat specialized displays for
explorative purposes, followed by model-based methods for
conditional independence models.

\subsection{Explorative Visualization Techniques}
\label{sec:flat}

@@

A first step in the explorative analysis of more complex tables is to
get a quick overview on the data. For that, 
all basic plots can be combined in pairwise displays,
arranged in a matrix similar to scatterplots in a pairs plot.
The diagonal cells contain the variable names, optionally with
univariate statistics, whereas the off-diagonal cells contain plots whose
variables are implicitly specified by the cells' position in the matrix. %'
In Figure~\ref{fig:pairs}, the diagonal cells show bar plots for the
distribution of the variables, and the off-diagonal cells mosaic plots
for the corresponding pair of variables. The plots suggest that
admission differs between male and female students, and between the departments,
and that the proportion of male and female students varies across the departments.

@@

%FIXME% Didn't we decide to use fewer plots in the paper and simply
%FIXME% point out that other combinations could also be used?
%FIXME% I would use for UCB:
%FIXME%   o explorative: highlighted doubledecker (currently Fig. 17)
%FIXME%   o model-based: conidtional shaded assoc (currently Fig. 20)
%FIXME% For Punishment:
%FIXME%   o explorative: highlighted mosaic( currently Fig. 21)
%FIXME%   o model-based: conditional shaded mosaic (currently Fig. 23)
%FIXME% The others should be omitted (I think) and the space gained should
%FIXME% be used for making the individual graphs larger. Preferably, we
%FIXME% should have the explorative and model-based plots for the same data
%FIXME% set on the same page, emphasizing that different aspects of the data
%FIXME% are brought out by the graphs.

The next step is to investigate three-way interactions among the variables.
In this example, we have a binary variable of interest (admission)
that shall be ``explained'' by the others. 
A natural way of representing such a three-way table is 
to draw a doubledecker plot nesting gender in
department and highlighting admission in the tiles (see Figure~\ref{fig:doubleucb}).
From the widths of the tiles it can be seen that an
unequal number of students apply for the six
departments, and from the highlighting we see that the admission rates differ as
well among the departments (roughly speaking, they decrease from A to F). The rates 
are equal for male and female students, except for department A where
\textit{more} female than male students are admitted.

@@

In addition to the explorative approaches 
to the visualization of multi-way tables outlined in 
the previous section, there are specialized displays 
designed for the visualization of conditional independence structures.

A first approach is to use pairs plots for model search:
we can use the position of the cells in the pairs matrix to select an
independence model and add a corresponding shading to the tiles to
visualize the residuals.
More precisely, each cell $a_{ij}$ in such a matrix defines two
variables $i$ and $j$ that can be used to specify the model visualized
in that cell. Typical hypotheses are: Variables $i$ and $j$ are
marginally independent; variables $i$ and $j$ are conditionally
independent, given all others; variables $i$ and $j$ are jointly
independent from all others.
% Figure~\ref{fig:pairs} shows a pairs display with mosaic plots visualizing mutual 
% independence in the upper triangle, 
% mosaics for \emph{joint} independence 
% in the lower triangle, and bar charts on the diagonal.
% In the upper part of the matrix, we can immediately 
% detect the independence of hair and gender, and eye and gender from the
% non-colored cells in the corresponding mosaic plots. On the other hand, hair and eye color 
% clearly are associated. In the lower part of the matrix, the mosaic
% at the intersection of hair and eye is the less colored one. Its shading
% represents the model of hair and eye color being jointly independent from
% gender, i.e., their association being homogeneous over the gender
% strata. It is the model that fits best, although still rejected at the
% 5\% level 
% ($p=\Sexpr{round(summary(loglm(~ Hair * Eye + Sex, data = HairEyeColor))$tests[2,3],4)}$): %$
% possibly, because too many brown-haired male
% students have blue eyes, and too few of them have brown eyes, than expected under
% the null model?

In our example, from the explorative analysis, we already suspect that
admission and gender are independent, given department, except for
department A. 
Therefore, a sensible way of representing the table is to 
nest admission into department instead of gender to get a stratified view.
The corresponding mosaic and association plots are shown
in Figures~\ref{fig:ucbmosaic}--\ref{fig:ucbassoc}. The shading visualizes the 
model of admission and gender being conditionally independent, given department. Clearly,
there is no gender bias in the departments except for department A: here, 
significantly \textit{more} female students are admitted then expected under independence.
Mosaic displays are already recognized as an excellent
visualization tool for log-linear models 
\cite{vcd:Friendly:1999,vcd:Theus+Lauer:1999,vcd:Hofmann:2001}.
Using the flat table representation, the association plot similarly extends from
the simple two-way case to a structured residual plot usable for 
the diagnostics of log-linear models. Especially if a
mosaic display contains very small or empty cells, the corresponding association plot
might provide an easier way to detect deviation patterns than the
residual-based shading added to the mosaic plot.

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}
\begin{center}
<<ucbmosaic,fig=TRUE,echo=FALSE,height=6,width=12,results=hide>>=
mosaic(aperm(UCBAdmissions, c(3,2,1)), data = UCBAdmissions, split = TRUE,
shade = TRUE, keep = FALSE, residuals = ucb_max$residuals,
gp_args = list(p.value = ucb_max$p.value), rep = c(Admit = FALSE)) 
@
\caption{Mosaic plot for the \data{UCB admissions} data.}
\label{fig:ucbmosaic}
\end{center}
\end{figure}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}
\begin{center}
<<ucbassoc,fig=TRUE,echo=FALSE,height=6,width=12,results=hide>>=
assoc(aperm(UCBAdmissions, c(3,2,1)), split = c(TRUE,FALSE,TRUE), shade = TRUE, 
residuals_type = "Pearson", residuals = ucb_max$residuals,
gp_args = list(p.value = ucb_max$p.value), rep = c(Admit = FALSE))
@
\caption{Association plot for the \data{UCB admissions} data.}
\label{fig:ucbassoc}
\end{center}
\end{figure}

@@

% Since mosaic displays are conditional plots by construction, we can use them
% for stratified data by using the conditioning variables as first splitting variables. 
% The mosaic in Fig.~\ref{fig:ucbmosaic} shows the distribution of admission
% and gender for all departments. A larger spacing for
% the department variable has been used to simplify the distinction between the 
% groups. 
As we have seen, we can use all plots for the analysis of stratified
data by using the conditioning variables first for splitting.
However, using a single plot, the strata are not corrected for the
marginal distribution of the stratifying variables. Especially for
very unevenly distributed marginals, the strata can become very
distorted, and the tiles, along with their shading, inscrutable.
An alternative is to use trellis-like layouts for visualizing \emph{partial}
tables for given strata, defined by the conditioning variables. 
All subtables are corrected for the marginals and thus, all
corresponding plots in the panels have the same size. 
In addition, trellis layouts help reduce the complexity of bigger tables. 
Figure~\ref{fig:ucbadmissions} 
visualizes the data by means of a conditional 
association plot. Each panel corresponds to a department, and contains
an association plot corresponding to the partial (fourfold) table of admission
and gender. Accordingly, the shading visualizes the
residuals from the corresponding conditional independence model (independence of gender
and admission, given department), stratified by department. Clearly,
the situation in department A (more women/less men admitted than
expected under the null hypothesis) causes the rejection of the
hypothesis of conditional independence. 
% The overall deviation pattern
% thus indicates that in contrast to the impression given by the aggregated table, male
% students were not advantaged, and that in department A,
% \emph{female} students were even more successful than their male colleagues.

@@

We first create a mosaic plot to get an overview of the data:
Attitude towards punishment clearly is the dependent variable 
here and should be used last for splitting. Furthermore, we might 
expect age and education to have stratifying influence on the results, so they should
be used first for splitting.
In fact, the question is whether memory has some influence on attitude, given 
age and education. We choose to create a three-way mosaic plot of the 
hypothesized explanatory variables (first splitting horizontically by age, 
then vertically by education, and finally horizontically again by memory), 
and to have attitude, the dependent variable, 
highlighted in the tiles (see Figure~\ref{fig:punishhigh}).
We can see that half of the people have more than 40 years of
age. Most people in this group have only completed elementary
school. For people over 25, a majority has completed secondary
school. 
%% FIXME: additional plot?
Among the younger people, fewer have memories of having been
punished as a child than among the older ones. 
%%
But surprisingly enough,
for people with elementary or secondary school education,
a higher proportion \emph{with} memories tends to
accept moderate punishment of children than those without: 
experienced violence seems to engender violence again.
However, for people under 40, the proportion with
people refusing punishment increases with the educational level.

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}
\begin{center}
<<punishhigh,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
tab <- xtabs(Freq ~ memory + attitude + age + education, data = Punishment)
mosaic(~ age + education + memory + attitude, data = tab, keep = FALSE,
       gp = gpar(fill = grey.colors(2)), spacing = spacing_highlighting, 
       rep = c(attitude = F), set_varnames = c(attitude = "attitude to punishment"),
       set_labels = list(age = c("15-24","25-39","40+")))
@
\caption{Mosaic plot with highlighting for the \data{punishment} data.}
\label{fig:punishhigh}
\end{center}
\end{figure}

So both age and education seem to matter here. But is there really any influence
of memory on the attitude towards 
punishment apart from that? Or, rephrased as a hypothesis:
are memory and attitude conditionally independent, given age and education?
We can visualize the model by means of a trellis layout, this time using partial mosaic
plots (see Figure~\ref{fig:punishcotab}) for the strata defined by all factorial 
combinations of the age and education levels. In each panel, we plot mosaic plots of 
memory and attitude, where again the explanatory variable (memory) is used first 
for splitting.
The shading tells us the model seems to be consistent with the data, 
except for people over 25
with elementary school education: under the null, more people should
have refused physical punishment of children than actually did.

@@

The advantage of the exploratory view 
is to visualize the joint distribution of all variables.
On the other hand, the visualization might be 
strongly influenced by the marginal distribution of education over age (in particular
    the large proportion of 40+ people with elementary education)
 which is not of interest in the conditional independence problem. The
 conditional mosaic plot suppresses the effect of all stratifying
 variables (and their interactions).

@@

Mosaic, association, and sieve plots can be used for multi-way tables
by applying them to flat representations. In addition, these basic
plots are leveraged by specialized frameworks 
for exploratory model search (pairs plots, highlighting, doubledecker plots) 
and stratified analysis (conditional plots).

>>>>

In the following, we will first treat specialized displays for
exploratory purposes, followed by model-based methods for
conditional independence models.

\subsection{Exploratory Visualization Techniques}
\label{sec:flat}

@@

A first step in the exploratory analysis of more complex tables is to
get a quick overview on the data. For that, 
all basic plots can be combined in pairwise displays,
arranged in a matrix similar to scatterplots in a pairs plot.
The diagonal cells contain the variable names, optionally with
univariate statistics, whereas the off-diagonal cells contain plots whose
variables are implicitly specified by the cells' position in the matrix. %'
In Figure~\ref{fig:pairs}, the diagonal cells show bar plots for the
distribution of the variables, and the off-diagonal cells mosaic plots
for the corresponding pair of variables. The plots suggest that
admission differs between male and female students, and between the departments,
and that the proportion of male and female students varies across the departments.
In particular, departments A and B have a higher proportion of male students
and a lower rejection rate compared to the other departments.

@@

The next step is to investigate three-way interactions among the variables.
In this example, we have a binary variable of interest (admission)
that shall be ``explained'' by the others. 
A natural way of representing such a three-way table is use a mosaic display
splitting first in the explanatory variables department and gender and highlighting
the resulting mosaic with respect to the dependent variable admissions. Here, we
use the vertical splits for both explanatory variables, resulting in the 
doubledecker plot in Figure~\ref{fig:doubleucb}. From the widths of the tiles it can be seen that an
unequal number of students apply for the six departments (with a particularly small
number of females in departments A and B),
and from the highlighting we see again that the admission rates differ
among the departments (roughly speaking, the admission rate is high for A and B,
low for F and in between for C to E). The rates 
are equal for male and female students, except for department A where
\textit{more} female than male students are admitted.

@@

In addition to the exploratory approaches 
to the visualization of multi-way tables outlined in 
the previous section, there are specialized displays 
designed for the visualization of conditional independence structures.

A first approach is to use pairs plots (such as that in Figure~\ref{fig:pairs}) for model search:
we can use the position of the cells in the pairs matrix to select an
independence model and add a corresponding shading to the tiles to
visualize the residuals.
More precisely, each cell $a_{ij}$ in such a matrix defines two
variables $i$ and $j$ that can be used to specify the model visualized
in that cell. Typical hypotheses are: Variables $i$ and $j$ are
marginally independent; variables $i$ and $j$ are conditionally
independent, given all others; variables $i$ and $j$ are jointly
independent from all others.
% Figure~\ref{fig:pairs} shows a pairs display with mosaic plots visualizing mutual 
% independence in the upper triangle, 
% mosaics for \emph{joint} independence 
% in the lower triangle, and bar charts on the diagonal.
% In the upper part of the matrix, we can immediately 
% detect the independence of hair and gender, and eye and gender from the
% non-colored cells in the corresponding mosaic plots. On the other hand, hair and eye color 
% clearly are associated. In the lower part of the matrix, the mosaic
% at the intersection of hair and eye is the less colored one. Its shading
% represents the model of hair and eye color being jointly independent from
% gender, i.e., their association being homogeneous over the gender
% strata. It is the model that fits best, although still rejected at the
% 5\% level 
% ($p=\Sexpr{round(summary(loglm(~ Hair * Eye + Sex, data = HairEyeColor))$tests[2,3],4)}$): %$
% possibly, because too many brown-haired male
% students have blue eyes, and too few of them have brown eyes, than expected under
% the null model?

Another approach is to visualize (the deviations from) a particular fitted model 
with similar ideas as in Section~\ref{sec:extensions}. Thus, mosaic displays can
be extended from purely explorative views to model-based views on the data by
residual-based shadings and association plots can directly visualize deviations
from a given model. Typically, the models considered are log-linear models, however
residual-based shadings are also conceivable for other types of models such as logistic
regression. %FIXME% mention this?
In our example, from the exploratory analysis, we already suspect that
admission and gender are independent, given department, except for
department A. 
Therefore, a sensible way of representing the table is to 
nest admission into department instead of gender to get a stratified view.
The corresponding mosaic and association plots are shown
in Figures~\ref{fig:ucbmosaic} and \ref{fig:ucbassoc}. The shading visualizes the 
model of admission and gender being conditionally independent, given department. Clearly,
there is no gender bias in the departments except for department A: here, 
significantly \textit{more} female students are admitted then expected under independence.
Mosaic displays are already recognized as an excellent
visualization tool for log-linear models 
\cite{vcd:Friendly:1999,vcd:Theus+Lauer:1999,vcd:Hofmann:2001}.
Using the flat table representation, the association plot similarly extends from
the simple two-way case to a structured residual plot usable for 
the diagnostics of log-linear models. Especially if a
mosaic display contains very small or empty cells, the corresponding association plot
might provide an easier way to detect deviation patterns than the
residual-based shading added to the mosaic plot.

\setkeys{Gin}{width=\textwidth}
\begin{figure}[t!]
\begin{center}
<<ucbmosaic,fig=TRUE,echo=FALSE,height=6,width=12,results=hide>>=
mosaic(aperm(UCBAdmissions, c(3,2,1)), data = UCBAdmissions, split = TRUE,
shade = TRUE, keep = FALSE, residuals = ucb_max$residuals,
gp_args = list(p.value = ucb_max$p.value), rep = c(Admit = FALSE)) 
@
\caption{Mosaic plot for the \data{UCB admissions} data.}
\label{fig:ucbmosaic}
\end{center}
\end{figure}

\setkeys{Gin}{width=\textwidth}
\begin{figure}[t!]
\begin{center}
<<ucbassoc,fig=TRUE,echo=FALSE,height=6,width=12,results=hide>>=
assoc(aperm(UCBAdmissions, c(3,2,1)), split = c(TRUE,FALSE,TRUE), shade = TRUE, 
residuals_type = "Pearson", residuals = ucb_max$residuals,
gp_args = list(p.value = ucb_max$p.value), rep = c(Admit = FALSE))
@
\caption{Association plot for the \data{UCB admissions} data.}
\label{fig:ucbassoc}
\end{center}
\end{figure}

@@

% Since mosaic displays are conditional plots by construction, we can use them
% for stratified data by using the conditioning variables as first splitting variables. 
% The mosaic in Fig.~\ref{fig:ucbmosaic} shows the distribution of admission
% and gender for all departments. A larger spacing for
% the department variable has been used to simplify the distinction between the 
% groups. 
As we have seen, we can use all plots for the analysis of stratified
data by using the conditioning variables first for splitting.
However, this also displays the marginal distribution of the conditioning variables
which might make it more difficult to interpret the dependent variables of
interest. Especially for
very unevenly distributed marginals, the strata can become very
distorted, and the tiles, along with their shading, inscrutable.
An alternative is to complement the conditioning in the model by
a conditioning in the plot, i.e., to use trellis-like layouts for
visualizing \emph{partial} tables as defined by the conditioning variables. 
All subtables are corrected for the marginals and thus, all
corresponding plots in the panels have the same size. 
In addition, trellis layouts help to reduce the complexity of bigger tables. 
Figure~\ref{fig:ucbadmissions} visualizes the data by means of a conditional 
association plot. Each panel corresponds to a department, and contains
an association plot corresponding to the partial (fourfold) table of admission
and gender. Accordingly, the shading visualizes the
residuals from the corresponding conditional independence model (independence of gender
and admission, given department), stratified by department. Clearly,
the situation in department A (more women/less men admitted than
expected under the null hypothesis) causes the rejection of the
hypothesis of conditional independence. In Figure~\ref{fig:ucbadmissions},
data-driven cut-offs (derived from the maximum test for conditional independence)
are used, resulting in a more colorful shading than the fixed cut-offs in
Figure~\ref{fig:ucbmosaic} and \ref{fig:ucbassoc}.

% The overall deviation pattern
% thus indicates that in contrast to the impression given by the aggregated table, male
% students were not advantaged, and that in department A,
% \emph{female} students were even more successful than their male colleagues.

@@

In a first step, we create a mosaic plot for an exploratory view on the data:
Attitude towards punishment clearly is the dependent variable 
here and should be used as the last split via highlighting. 
Furthermore, we expect age and education to have an influence on both the memory
and the attitude, hence they should be used first for splitting as stratifying
variables. In fact, the question is whether memory has some influence on attitude, given 
age and education. We choose to create a three-way mosaic plot of the 
hypothesized explanatory variables (first splitting horizontically by age, 
then vertically by education, and finally horizontically again by memory), 
and to have attitude, the dependent variable, 
highlighted in the tiles (see Figure~\ref{fig:punishhigh}).
We can see that half of the people have more than 40 years of
age, most of which of only completed elementary school (much more than 
in the other age groups). Furthermore, it can be seen that the proportion
of people with memories of punishments increases with age and that
the approval rate decreases with education. As for the question whether
attitude depends on memory, the plot suggests quite clearly (and somewhat
surprisingly) that for those people with elementary school education,
a higher proportion \emph{with} memories tends to
accept moderate punishment of children than those without: 
experienced violence seems to engender violence again.
For the other education groups, the picture is less clear: some cells 
indicate the same association while others do not.

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}
\begin{center}
<<punishhigh,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
tab <- xtabs(Freq ~ memory + attitude + age + education, data = Punishment)
dimnames(tab)[[3]][3] <- "40+"
mosaic(~ age + education + memory + attitude, data = tab, keep = FALSE,
       gp = gpar(fill = grey.colors(2)), spacing = spacing_highlighting, 
       rep = c(attitude = FALSE))
@
\caption{Mosaic plot with highlighting for the \data{punishment} data.}
\label{fig:punishhigh}
\end{center}
\end{figure}

In summary, both age and education clearly matter here.
But is there really any significant influence of memory on the attitude towards 
punishment apart from that? Or, rephrased as a hypothesis:
Are memory and attitude conditionally independent, given age and education?
To focus on this question, we will take another model-based view on the data.
To reduce the complexity of the information to the question of interest, we employ
partial mosaic plots in a trellis layout (conditioning on age and education) with
residual-based shading and data-driven cut-offs (see Figure~\ref{fig:punishcotab}).
This visualization brings out much more clearly that the increased approval rate
among people with memories of punishments is present mainly in the elementary education
column and, to a lesser degree, in the age 40+ row. Moreover, this association is
only significant in two cells (the two older age groups with elementary education)
where fewer people without memories do not approve punishment than expected under
(conditional) independence.

@@

The advantage of the exploratory view 
is to visualize the joint distribution of all variables.
On the other hand, the visualization might be 
strongly influenced by the marginal distribution of education over age (in particular
the large proportion of 40+ people with elementary education)
which is not of interest in the conditional independence problem. The
partial mosaic plot suppresses this effect by complementing the 
conditioning in the model with conditioning in the visualization.

@@

Mosaic, association, and sieve plots can be used for visualizing multi-way tables
by applying them to flat representations. Mosaic plots are particularly useful
for exploratory analysis whereas the other two require specification of a certain
independence model for which deviations should be brought out. In addition, specialized
`flavours' of the plots can leverage the
exploratory (pairs plots, highlighting, doubledecker plots) 
or model-based analysis (residual-based shading, conditional plots).
